---
title: "R Notebook"
output: html_notebook
---

```{r}
## Data sets are downloaded from the Data Expo '09 web site at
## http://stat-computing.org/dataexpo/2009/the-data.html
for (year in 1987:2008) {
    filename <- paste(year, "csv.bz2", sep = ".")
    if ( !file.exists(filename) ) {
        url <- paste("http://stat-computing.org/dataexpo/2009/",
                          year, ".csv.bz2", sep = "")
        cat("Downloading data file ", filename, "\n", sep = "")
        download.file(url, filename)
    }
}
```

```{r}
## Read sample file to get column names and types
d <- read.csv("2008.csv.bz2")
integer.columns <- sapply(d, is.integer)
factor.columns  <- sapply(d, is.factor)
factor.levels   <- lapply(d[, factor.columns], levels)
n.rows <- 0L
```

```{r}
## Process each file determining the factor levels
## TODO: Combine with next loop
for (year in 1987:2008) {
    filename <- paste(year, "csv.bz2", sep = ".")
    cat("Processing ", filename, "\n", sep = "")
    d <- read.csv(filename)
    n.rows <- n.rows + NROW(d)
    new.levels <- lapply(d[, factor.columns], levels)
    for ( i in seq(1, length(factor.levels)) ) {
        factor.levels[[i]] <- c(factor.levels[[i]], new.levels[[i]])
    }
    rm(d)
}
save(integer.columns, factor.columns, factor.levels, file = "factors.RData")
```

```{r}
## Now convert all factors to integers so we can create a bigmatrix of the data
col.classes <- rep("integer", length(integer.columns))
col.classes[factor.columns] <- "character"
cols  <- which(factor.columns)
first <- TRUE
csv.file <- "airlines.csv"   # Write combined integer-only data to this file
csv.con  <- file(csv.file, open = "w")

for (year in 1987:2008) {
    file.name <- paste(year, "csv.bz2", sep = ".")
    cat("Processing ", file.name, "\n", sep = "")
    d <- read.csv(file.name, colClasses = col.classes)
    ## Convert the strings to integers
    for ( i in seq(1, length(factor.levels)) ) {
        col <- cols[i]
        d[, col] <- match(d[, col], factor.levels[[i]])
    }
    write.table(d, file = csv.con, sep = ",", 
                row.names = FALSE, col.names = first)
    first <- FALSE
}
close(csv.con)
```

```{bash}
echo New format:
head /home/mint/Documents/AirlineR/airlines.csv
echo
echo Original format:
bzcat /home/mint/Documents/AirlineR/1987.csv.bz2 | head -10
```
```{bash}
echo New format:
tail /home/mint/Documents/AirlineR/airlines.csv
echo
echo Original format:
bzcat /home/mint/Documents/AirlineR/2008.csv.bz2 | tail -10
```

```{r}
## Now convert to a big.matrix
library("bigmemory")
# backing.file    <- "airlines.bin"
# descriptor.file <- "airlines.des"
data <- read.big.matrix("airlines.csv", header = TRUE,
                        type = "integer",
                        backingfile = "airlines.bin",
                        descriptorfile = "airlines.des",
                        extraCols = c("age", "Hour"))
```

```{r}
#matches eric
print(dim(data))
```

```{bash}
# -rw-r--r--@ 1 eclarson  staff  11785518426 Jan 14 17:06 Data/AirlineDataAll.csv
# the size of the csv file is a little different due to the different int conversion
ls -all /home/mint/Documents/AirlineR/airlines.csv
```

```{bash}
# -rw-r--r--@ 1 eclarson  staff  14824196281 Jan 15 12:08 Data/airline.bin
# the resulting bin file is only one byte off - probably not significant
ls -all /home/mint/Documents/AirlineR/airlines.bin
```

```{r}
library("parallel")
## Use parallel processing if available
## (Multicore is for "anything-but-Windows" platforms)
if ( require("parallel") ) {
    library("doMC")
    registerDoMC(cores=3)
} else {
    warning("Consider registering a multi-core 'foreach' processor.")
}
```

```{r}
library("bigmemory")
library("bigtabulate")


## attach the binary backing file through its descriptor
descriptor.file <- "airlines.des"
#data <- attach.big.matrix(dget(descriptor.file))
data <- attach.big.matrix(descriptor.file)
```

```{r}
origin_indicies <- bigsplit(data, "Origin", splitcol = NA_real_)
```

```{r}
# lets do some basics by graphing aggregated data
# not sure why not working
counts <- sapply(origin_indicies , function(i) length(i))
```

```{r}
# Analyzing Age of Plane
acindices <- bigsplit(data, 'TailNum', splitcol = NA_real_)
```

```{r}
# define a 'birthplace' function
# we use the very first recorded flight for a specific plane as an estimate of its age

# let's first just define e function to get the 
birthmonth <- function(y) {
    # assume that the input is one matric of values from one plane
    
    # get minimum year for this plane
    minYear <- min(y[,'Year'], na.rm=TRUE)
    # get a subset of the dataset for only this minimum year
    these <- which(y[,'Year']==minYear)
    # get the minimum month from the years
    minMonth <- min(y[these,'Month'], na.rm=TRUE) 
    # now just return the number of months since 00 AD 
    return(12*minYear + minMonth - 1)
}
```

```{r}
# recall that sapply will send the different indices into the given function
# here, we send in the indices of the plane, with the year and month
# this will be done for every plane based upon the acindices grouping
acStart <- sapply(acindices , function(i) birthmonth(data[i, c('Year','Month'), drop= FALSE]))
```

```{r}
#of course, we really wanted this to be a fast computation 
# and we could split the operations using a foreach parallel loop

# setup the parallel for package (done above)
# library(doMC)
# registerDoMC(cores=3)
```

```{r}
# now run it--in this example we only get e mild speedup because the computation is 
# fairly quick and there are a lot of groups to be made

# note that this package handles a bit more gracefully the attached memory mapped files to the worker threads
# this works similarly to the parallel for loop from your book, but slightly simpler
acStart <- foreach(i=acindices , .combine=c)%dopar% {
    return(birthmonth(data[i, c('Year', 'Month'), drop=FALSE]))
}
```

(Python code kept for reference for now)  
Which plane is the youngest?  
idx = np.argmax(acStart)  

print 'The youngest plane is', list(unique_values['TailNum'])[idx],  
print 'and flew starting in the year %.0f'%(acStart[idx]/12.0)  

The youngest plane is N5WCAA and flew starting in the year 2009

```{r}
# now lets save the age of th plane at the time it flew a given route
data[,"age"] <- data[,"Year"]*12+data[,"Month"]-acStart[data[,"TailNum"]]
```

```{r}
# save the scheduled hour of departure

data[,"Hour"] <- (data[,"CRSDepTime"] / 100)

print(data[,"Hour"])
```

```{r}
library("biganalytics")

blm <- biglm.big.matrix( ArrDelay ~ age, data=data)
```

```{r}
summary(blm)
```

COMPARED TO ERIC RESULTS - Notice the smaller sample size  
Large data regression model: biglm(formula = formula, data = data, ...)  
Sample size =  120947440  
              Coef   (95%    CI)     SE p  
(Intercept) 7.1203 7.1146 7.1260 0.0028 0  
age         0.0045 0.0044 0.0046 0.0000 0  

```{r}
## Plot Arrival Dealy by Hour
arrive_delay_hour <- bigtabulate(data,
                   ccols = "Hour",
                   summary.cols = "ArrDelay", summary.na.rm = TRUE, splitcol = NA_real_)
```

```{r}
hour.names <- c(0, 1, 2, 3, 4, 5, 6, 7,
                8, 9, 10, 11, 12, 13, 14,
                15, 16, 17, 18, 19, 20, 21,
                22, 23, 24)

stat.names <- dimnames(arrive_delay_hour$summary[[1]])[2][[1]]

arrive_delay_hour.p <- cbind(matrix(unlist(arrive_delay_hour$summary), byrow = TRUE,
                      nrow = length(arrive_delay_hour$summary),
                      ncol = length(stat.names),
                      dimnames = list(hour.names, stat.names)),
               ValidObs = arrive_delay_hour$table)

print(arrive_delay_hour.p)
```

```{r}
plot(arrive_delay_hour.p[, "mean"], type = "l", ylab="Average arrival delay")
```

```{r}
## Plot Arrival Dealy by Day
arrive_delay_day <- bigtabulate(data,
                   ccols = "DayOfWeek",
                   summary.cols = "ArrDelay", summary.na.rm = TRUE, splitcol = NA_real_)
```

```{r}
day.names <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday",
               "Saturday", "Sunday")

## matches output from http://www.sas.rochester.edu/psc/thestarlab/help/Big-Data-WP.pdf
stat.names <- dimnames(arrive_delay_day$summary[[1]])[2][[1]]
arrive_delay_day.p <- cbind(matrix(unlist(arrive_delay_day$summary), byrow = TRUE,
                      nrow = length(arrive_delay_day$summary),
                      ncol = length(stat.names),
                      dimnames = list(day.names, stat.names)),
               ValidObs = arrive_delay_day$table)
print(arrive_delay_day.p)
```

```{r}
## Figure 1 - X axis are days of the week - nice if it would say so
## matches output from http://www.sas.rochester.edu/psc/thestarlab/help/Big-Data-WP.pdf
plot(arrive_delay_day.p[, "mean"], type = "l", ylab="Average arrival delay")
```

```{r}
## Plot Arrival Dealy by Month
arrive_delay_month = bigtabulate(data,
                   ccols = "Month",
                   summary.cols = "ArrDelay", summary.na.rm = TRUE, splitcol = NA_real_)
```

```{r}
month.names = c("Jan", "Feb", "Mar", "Apr", "May",
               "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

stat.names <- dimnames(arrive_delay_month$summary[[1]])[2][[1]]

arrive_delay_month.p <- cbind(matrix(unlist(arrive_delay_month$summary), byrow = TRUE,
                      nrow = length(arrive_delay_month$summary),
                      ncol = length(stat.names),
                      dimnames = list(month.names, stat.names)),
               ValidObs = arrive_delay_month$table)

print(arrive_delay_month.p)
```

```{r}
plot(arrive_delay_month.p[, "mean"], type = "l", ylab="Average arrival delay")
```

```{r}
## Plot Arrival Dealy by Year
arrive_delay_year = bigtabulate(data,
                   ccols = "Year",
                   summary.cols = "ArrDelay", summary.na.rm = TRUE, splitcol = NA_real_)
```

```{r}
year.names = c(1987, 1988, 1989, 1990, 1991, 1992, 1993,
                1994, 1995, 1996, 1997, 1998, 1999, 2000,
                2001, 2002, 2003, 2004, 2005, 2006, 2007, 2008)

stat.names <- dimnames(arrive_delay_year$summary[[1]])[2][[1]]

arrive_delay_year.p <- cbind(matrix(unlist(arrive_delay_year$summary), byrow = TRUE,
                      nrow = length(arrive_delay_year$summary),
                      ncol = length(stat.names),
                      dimnames = list(year.names, stat.names)),
               ValidObs = arrive_delay_year$table)

print(arrive_delay_year.p)
```
```{r}
plot(arrive_delay_year.p[, "mean"], type = "l", ylab="Average arrival delay")
```

### Which airports are most likely to be delayed flying into or out of?
```{r}
pct_delayed_by_origin <- bigtabulate(data,
                                     ccols = "Origin",
                                     summary.cols = ("DepDelay">15), summary.na.rm = TRUE, splitcol = NA_real_)

pct_late_by_dest <- bigtabulate(data,
                                     ccols = "Dest",
                                     summary.cols = ("ArrDelay">15), summary.na.rm = TRUE, splitcol = NA_real_)

pct_delayed_by_origin_dest <- bigtabulate(data,
                                     ccols = c("Origin", "Dest"),
                                     summary.cols = ("DepDelay">15), summary.na.rm = TRUE, splitcol = NA_real_)

pct_late_by_origin_dest <- bigtabulate(data,
                                     ccols = c("Origin", "Dest"),
                                     summary.cols = ("ArrDelay">15), summary.na.rm = TRUE, splitcol = NA_real_)

# stat.names <- dimnames(pct_delayed_by_origin$summary[[1]])[2][[1]]
# 
# pct_delayed_by_origin <- cbind(matrix(unlist(pct_delayed_by_origin$summary), byrow = TRUE,
#                       nrow = length(pct_delayed_by_origin$summary),
#                       ncol = length(stat.names),
#                       dimnames = list(origin_indicies, stat.names)),
#                ValidObs = pct_delayed_by_origin$table)
# 
# head(pct_delayed_by_origin)
```
```{r}
rm(male_data, paramGrid, run_data, birth1, birthNum, btime, bTimes, bTimes1, counts, cTimes, cTimes1, d0, dtime)
```

