---
title: "R Notebook"
output: html_notebook
---

```{r}
## Data sets are downloaded from the Data Expo '09 web site at
## http://stat-computing.org/dataexpo/2009/the-data.html
for (year in 1987:2008) {
    filename <- paste(year, "csv.bz2", sep = ".")
    if ( !file.exists(filename) ) {
        url <- paste("http://stat-computing.org/dataexpo/2009/",
                          year, ".csv.bz2", sep = "")
        cat("Downloading data file ", filename, "\n", sep = "")
        download.file(url, filename)
    }
}
```

```{r}
## Read sample file to get column names and types
d <- read.csv("2008.csv.bz2")
integer.columns <- sapply(d, is.integer)
factor.columns  <- sapply(d, is.factor)
factor.levels   <- lapply(d[, factor.columns], levels)
n.rows <- 0L
```

```{r}
## Process each file determining the factor levels
## TODO: Combine with next loop
for (year in 1987:2008) {
    filename <- paste(year, "csv.bz2", sep = ".")
    cat("Processing ", filename, "\n", sep = "")
    d <- read.csv(filename)
    n.rows <- n.rows + NROW(d)
    new.levels <- lapply(d[, factor.columns], levels)
    for ( i in seq(1, length(factor.levels)) ) {
        factor.levels[[i]] <- c(factor.levels[[i]], new.levels[[i]])
    }
    rm(d)
}
save(integer.columns, factor.columns, factor.levels, file = "factors.RData")
```

```{r}
## Now convert all factors to integers so we can create a bigmatrix of the data
col.classes <- rep("integer", length(integer.columns))
col.classes[factor.columns] <- "character"
cols  <- which(factor.columns)
first <- TRUE
csv.file <- "airlines.csv"   # Write combined integer-only data to this file
csv.con  <- file(csv.file, open = "w")

for (year in 1987:2008) {
    file.name <- paste(year, "csv.bz2", sep = ".")
    cat("Processing ", file.name, "\n", sep = "")
    d <- read.csv(file.name, colClasses = col.classes)
    ## Convert the strings to integers
    for ( i in seq(1, length(factor.levels)) ) {
        col <- cols[i]
        d[, col] <- match(d[, col], factor.levels[[i]])
    }
    write.table(d, file = csv.con, sep = ",", 
                row.names = FALSE, col.names = first)
    first <- FALSE
}
close(csv.con)
```

```{bash}
echo New format:
head /home/mint/Documents/AirlineR/airlines.csv
echo
echo Original format:
bzcat /home/mint/Documents/AirlineR/1987.csv.bz2 | head -10
```
```{bash}
echo New format:
tail /home/mint/Documents/AirlineR/airlines.csv
echo
echo Original format:
bzcat /home/mint/Documents/AirlineR/2008.csv.bz2 | tail -10
```

```{r}
## Now convert to a big.matrix
library("bigmemory")
# backing.file    <- "airlines.bin"
# descriptor.file <- "airlines.des"
data <- read.big.matrix("airlines.csv", header = TRUE,
                        type = "integer",
                        backingfile = "airlines.bin",
                        descriptorfile = "airlines.des",
                        extraCols = c("age", "Hour", "Hour_scaled", "Distance_scaled", "Age_scaled"))
```

```{r}
#matches eric
print(dim(data))
```

```{bash}
# -rw-r--r--@ 1 eclarson  staff  11785518426 Jan 14 17:06 Data/AirlineDataAll.csv
# the size of the csv file is a little different due to the different int conversion
ls -all /home/mint/Documents/AirlineR/airlines.csv
```

```{bash}
# -rw-r--r--@ 1 eclarson  staff  14824196281 Jan 15 12:08 Data/airline.bin
# the resulting bin file is only one byte off - probably not significant
ls -all /home/mint/Documents/AirlineR/airlines.bin
```

```{r}
library("parallel")
## Use parallel processing if available
## (Multicore is for "anything-but-Windows" platforms)
if ( require("parallel") ) {
    library("doMC")
    registerDoMC(cores=3)
} else {
    warning("Consider registering a multi-core 'foreach' processor.")
}
```

```{r}
library("bigmemory")
library("bigtabulate")


## attach the binary backing file through its descriptor
descriptor.file <- "airlines.des"
#data <- attach.big.matrix(dget(descriptor.file))
data <- attach.big.matrix(descriptor.file)
```

```{r}
origin_indicies <- bigsplit(data, "Origin", splitcol = NA_real_)
```

```{r}
# lets do some basics by graphing aggregated data
# not sure why not working
counts <- sapply(origin_indicies , function(i) length(i))
```

```{r}
# Analyzing Age of Plane
acindices <- bigsplit(data, 'TailNum', splitcol = NA_real_)
```

```{r}
# define a 'birthplace' function
# we use the very first recorded flight for a specific plane as an estimate of its age

# let's first just define e function to get the 
birthmonth <- function(y) {
    # assume that the input is one matric of values from one plane
    
    # get minimum year for this plane
    minYear <- min(y[,'Year'], na.rm=TRUE)
    # get a subset of the dataset for only this minimum year
    these <- which(y[,'Year']==minYear)
    # get the minimum month from the years
    minMonth <- min(y[these,'Month'], na.rm=TRUE) 
    # now just return the number of months since 00 AD 
    return(12*minYear + minMonth - 1)
}
```

```{r}
# recall that sapply will send the different indices into the given function
# here, we send in the indices of the plane, with the year and month
# this will be done for every plane based upon the acindices grouping
acStart <- sapply(acindices , function(i) birthmonth(data[i, c('Year','Month'), drop= FALSE]))
```

```{r}
#of course, we really wanted this to be a fast computation 
# and we could split the operations using a foreach parallel loop

# setup the parallel for package (done above)
# library(doMC)
# registerDoMC(cores=3)
```

## Feature Preparation and Creation

In preparing the files for analysis, we also wanted to create new features that could be used throughout the rest of the processing. There are 2 features created for the regression with 2 additional features as intermediaries. The first feature is the Hour variable. As visualized in a following section, later hours have more delays. It is transformed from the scheduled departure time (CRSDepTime) to the hour of the day.

The other feature created is the estimated plane age. The age could be a potential factor in delays and will be used during the regression analysis of the data. The age of a plane is estimated from the day of the first flight found in the data. It is calculated by determining the number of months from 0 A.D. for each flight, and the first flight of each plane. The difference results in the number of months between the first flight and each flight, or the estimated plane age in months.

```{r}
# now run it--in this example we only get e mild speedup because the computation is 
# fairly quick and there are a lot of groups to be made

# note that this package handles a bit more gracefully the attached memory mapped files to the worker threads
# this works similarly to the parallel for loop from your book, but slightly simpler
acStart <- foreach(i=acindices , .combine=c)%dopar% {
    return(birthmonth(data[i, c('Year', 'Month'), drop=FALSE]))
}
```

(Python code kept for reference for now)  
Which plane is the youngest?  
idx = np.argmax(acStart)  

print 'The youngest plane is', list(unique_values['TailNum'])[idx],  
print 'and flew starting in the year %.0f'%(acStart[idx]/12.0)  

The youngest plane is N5WCAA and flew starting in the year 2009

```{r}
# now lets save the age of th plane at the time it flew a given route
data[,"age"] <- data[,"Year"]*12+data[,"Month"]-acStart[data[,"TailNum"]]
```

```{r}
# save the scheduled hour of departure

data[,"Hour"] <- (data[,"CRSDepTime"] / 100)

print(data[,"Hour"])
```

## Flight Delays

According to the FAA, when a schedule airflight departs more than 15 minutes after its scheduled time, it is considered officially delayed. We utilize the same logic for arrival times to determine if a flight is arriving late and is therefore delayed by arrival rather than by departure. Only departures and arrivals 15 minutes past the scheduled time will be considered late in the analysis.

http://aspmhelp.faa.gov/index.php/Types_of_Delay

## Visualization of Average Delay

We first take a look at the average delay for flights from the data set. Below, we look at the different representations for the average delay time for year, month, day, hour, and finally by carrier.

An interesting plot to observe is the average delay by hour. It seems that there is a constant increase in delay times as the day gets going. From 6am to around 8pm, there is a constant increase in the average delay time. It starts less than 2 minutes on average for delays and increases all the way up to nearly 14 minutes. After 8pm, the time decreases again into the early morning, leveling back to "low" levels less than 4 minutes around 2am.

```{r}
## Plot Arrival Dealy by Hour
arrive_delay_hour <- bigtabulate(data,
                   ccols = "Hour",
                   summary.cols = "ArrDelay", summary.na.rm = TRUE, splitcol = NA_real_)
```

```{r}
hour.names <- c(0, 1, 2, 3, 4, 5, 6, 7,
                8, 9, 10, 11, 12, 13, 14,
                15, 16, 17, 18, 19, 20, 21,
                22, 23, 24)

stat.names <- dimnames(arrive_delay_hour$summary[[1]])[2][[1]]

arrive_delay_hour.p <- cbind(matrix(unlist(arrive_delay_hour$summary), byrow = TRUE,
                      nrow = length(arrive_delay_hour$summary),
                      ncol = length(stat.names),
                      dimnames = list(hour.names, stat.names)),
               ValidObs = arrive_delay_hour$table)

print(arrive_delay_hour.p)
```

```{r}
plot(arrive_delay_hour.p[, "mean"], type = "l", ylab="Average arrival delay")
```

The average delay by day of the week does not have much to note. Delays are fairly stable around 8 minutes on average for each day, with day 5, Friday, being higher at 10 minutes. This is most likely due to high travel frequency on that day.

```{r}
## Plot Arrival Dealy by Day
arrive_delay_day <- bigtabulate(data,
                   ccols = "DayOfWeek",
                   summary.cols = "ArrDelay", summary.na.rm = TRUE, splitcol = NA_real_)
```

```{r}
day.names <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday",
               "Saturday", "Sunday")

## matches output from http://www.sas.rochester.edu/psc/thestarlab/help/Big-Data-WP.pdf
stat.names <- dimnames(arrive_delay_day$summary[[1]])[2][[1]]
arrive_delay_day.p <- cbind(matrix(unlist(arrive_delay_day$summary), byrow = TRUE,
                      nrow = length(arrive_delay_day$summary),
                      ncol = length(stat.names),
                      dimnames = list(day.names, stat.names)),
               ValidObs = arrive_delay_day$table)
print(arrive_delay_day.p)
```

```{r}
## Figure 1 - X axis are days of the week - nice if it would say so
## matches output from http://www.sas.rochester.edu/psc/thestarlab/help/Big-Data-WP.pdf
plot(arrive_delay_day.p[, "mean"], type = "l", ylab="Average arrival delay")
```

When we look at the average delay by month, we see December having the largest delay times with an average delay time of nearly 12 minutes. The month of September is the lowest month of delays. These highs and lows are most likely attributed to the flight patterns of people. Many people do not travel in September because school is starting back up and people have already taken time off for the summer. There are also a lot more travel conducted during December because of Christmas, one of the more popular holidays for travel.

```{r}
## Plot Arrival Dealy by Month
arrive_delay_month = bigtabulate(data,
                   ccols = "Month",
                   summary.cols = "ArrDelay", summary.na.rm = TRUE, splitcol = NA_real_)
```

```{r}
month.names = c("Jan", "Feb", "Mar", "Apr", "May",
               "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

stat.names <- dimnames(arrive_delay_month$summary[[1]])[2][[1]]

arrive_delay_month.p <- cbind(matrix(unlist(arrive_delay_month$summary), byrow = TRUE,
                      nrow = length(arrive_delay_month$summary),
                      ncol = length(stat.names),
                      dimnames = list(month.names, stat.names)),
               ValidObs = arrive_delay_month$table)

print(arrive_delay_month.p)
```

```{r}
plot(arrive_delay_month.p[, "mean"], type = "l", ylab="Average arrival delay")
```

For the average delay by year, there dont seem to be many deviations. There are two significant years higher than 10 minutes in 2000 and 2007. We aloso see several years with low average delays less than 6 minutes including 1991, 1992, 2002, and 2003.

```{r}
## Plot Arrival Dealy by Year
arrive_delay_year = bigtabulate(data,
                   ccols = "Year",
                   summary.cols = "ArrDelay", summary.na.rm = TRUE, splitcol = NA_real_)
```

```{r}
year.names = c(1987, 1988, 1989, 1990, 1991, 1992, 1993,
                1994, 1995, 1996, 1997, 1998, 1999, 2000,
                2001, 2002, 2003, 2004, 2005, 2006, 2007, 2008)

stat.names <- dimnames(arrive_delay_year$summary[[1]])[2][[1]]

arrive_delay_year.p <- cbind(matrix(unlist(arrive_delay_year$summary), byrow = TRUE,
                      nrow = length(arrive_delay_year$summary),
                      ncol = length(stat.names),
                      dimnames = list(year.names, stat.names)),
               ValidObs = arrive_delay_year$table)

print(arrive_delay_year.p)
```
```{r}
plot(arrive_delay_year.p[, "mean"], type = "l", ylab="Average arrival delay")
```

### Which airports are most likely to be delayed flying into or out of?

The analysis conducted below looks at the combination of departure delays and arrival delays for each airport in the dataset. We observe the average time of delay for this analysis. The higher the average delay time, regardless of number of flights, the more likely it is that a flight will be delayed.

From our first analysis, we see that there are 4 airports that have an average departure delay time over 100 minutes. This includes the largest average delay for departures of 203 minutes. It is likely this airport has a small number of flights to obtain this average, but they have a minimum delay of 122 minutes and a maximum of 249 minutes.

```{r}
# Finds delays based on Origin
pct_delayed_by_origin <- bigtabulate(data,
                                     ccols = "Origin",
                                     summary.cols = ("DepDelay"), summary.na.rm = TRUE, splitcol = NA_real_)
```
```{r}
# Orders summary data by mean (3rd value in list)
pct_delayed_by_origin_top = rev(pct_delayed_by_origin$summary[order(sapply(pct_delayed_by_origin$summary, '[[', i=3))])

# Prints top 5 airports with a departure delay
head(pct_delayed_by_origin_top$summary)
```

We also analyzed the delay in arrivals at destinations.

```{r}
# Finds delayed Arrivals based on Destination
pct_late_by_dest <- bigtabulate(data,
                                    ccols = "Dest",
                                     summary.cols = ("ArrDelay"), summary.na.rm = TRUE, splitcol = NA_real_)
```

```{r}
# Orders summary data by mean (3rd value in list)
pct_late_by_dest_top = pct_late_by_dest$summary[order(sapply(pct_late_by_dest$summary, '[[', i=3))]

# Prints top 5 airports with an arrival delay
head(pct_late_by_dest_top)
```

### Which flights with same origin and destiation are most likely to be delayed

The next question we want to answer are which typical flights are more likely to be delayed. Using a similar process as conducted previously, we look at the average time of departure delays and arrival delays and average the value to get the total value of delayed flights. For this question, however, we want all flights where the origin and destination are the same instead of looking at individual airports.



```{r}
# Gets delayed departure flights by Origin and Destination
pct_delayed_by_origin_dest <- bigtabulate(data,
                                     ccols = c("Origin", "Dest"),
                                     summary.cols = ("DepDelay"), summary.na.rm = TRUE, splitcol = NA_real_)
```

```{r}
# Orders summary data by mean (3rd value in list)
pct_delayed_by_origin_dest_top <- pct_delayed_by_origin_dest$summary[order(sapply(pct_delayed_by_origin_dest$summary, '[[', i=3))]

# Prints top 5 flights with a departure delay
head(pct_delayed_by_origin_dest_top)
```

```{r}
# Gets delayed arrival flights by Origin and Destination
pct_late_by_origin_dest <- bigtabulate(data,
                                     ccols = c("Origin", "Dest"),
                                     summary.cols = ("ArrDelay"), summary.na.rm = TRUE, splitcol = NA_real_)
```

```{r}
# Orders summary data by mean (3rd value in list)
pct_late_by_origin_dest_top <- pct_late_by_origin_dest$summary[order(sapply(pct_late_by_origin_dest$summary, '[[', i=3))]

# Prints top 5 flights with an arrival delay
head(pct_late_by_origin_dest_top)
```

### Regression

The following features will be explore to predict if the flight will have departure delay

The predicted variable will be:
- Departure Delay (DepDelay)

The explanatory variables:
- Scheduled departure hour (Hour)
- Flight distance (Distance)
- Age of plane (Age)

```{r}
library("biganalytics")
blm <- biglm.big.matrix( ArrDelay ~ age, data=data)
```

```{r}
summary(blm)
```


COMPARED TO ERIC RESULTS - Notice the smaller sample size  
Large data regression model: biglm(formula = formula, data = data, ...)  
Sample size =  120947440  
              Coef   (95%    CI)     SE p  
(Intercept) 7.1203 7.1146 7.1260 0.0028 0  
age         0.0045 0.0044 0.0046 0.0000 0 


## Regression Results

We conducted regression on the data set and wanted to determine what variables may affect the departure delay time the most. Due to the size of the data set, we selected variables based on previous observations and visualizations. The main variables we wanted to observe were the hour the flight was supposed to take off and the distance it was to travel. We conducted this regression twice. The first run was prior to scaling the data to one another while the second took this scaling into account. Scaling the data reveals comparable levels of importance.

In our first run, we saw that hour was the dominant factor of these two values in predicting the time of delay for a flight. The hour variable had an average coefficient of 0.7263 (±.001) compared to the 0.0015 (±0) value of distance.

This is a skewed value however because the hour value ranges from 0 to 23 and the distance value can be quite large depending on the flight being conducted. To observe this relationship appropriately, we scale the hour and distance values. After doing this scaling, we see coefficients of 3.453 (±.01) for the departure hour and 0.82 (±.06) for the distance. This means that the delay time for departure can be calculated with a factor of 3.453 multiplied by the scaled hour and a factor of 0.82 multiplied by the scaled flight distance. The Mean Squared Error was 866.89, slightly larger than the non-scaled analysis of 804.55, with an R Squared value of -0.067.

```{r}
linReg_hourDist <- biglm.big.matrix( DepDelay ~ Hour+Distance, data=data)
```

```{r}
summary(linReg_hourDist)
```

Our second analysis added the age of the plane to the analysis. The age of the plane was represented in number of months and was calculated using the difference between the first recorded flight and the flight being conducted. In this analysis, we continued using the hour of departure and distance of the scheduled flight. We saw coefficients of 0.8424 (±.0015) for the hour of departure for the flight, 0.0010 (±.0001) for the distance of the flight, and 0.0001 (±.0001) for the age of the plane. Age confidence interval includes zero, therefore, the evidence points this variable is not a significant predictor. The hour plays the largest role in affecting the delay time when we use these coefficients. The Mean Squared Error of this regression for 40% of the data was 866.01, which is just slightly better than the initial regression using just hour of departure and distance of flight.


```{r}
linReg_hourDistAge <- biglm.big.matrix( DepDelay ~ age+Hour+Distance, data=data)
```

```{r}
summary(linReg_hourDistAge)
```

## Comparison between Python and R Analysis

